---
title: "Customization guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customization guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  dev = "png",
  out.width = "100%",
  fig.retina = 2
)
```

```{r, message=FALSE}
library(vistool)
library(mlr3)
library(mlr3learners)
set.seed(1)
```

The vistool package provides a flexible theme system that separates **what to plot** from **how to plot**.
The three-step workflow:

1. **Initialize** the Visualizer using `as_visualizer()` (computational setup only - the **what**) and set the theme,
2. **Add** visualization layers with automatic color management,
3. **Plot** using `plot()` with optional customization, and optionally `save()`,

supports consistent customization across all visualization types with a clear precedence hierarchy.

## Theme system overview

### Theme precedence (highest to lowest)

1. **Layer-specific style**: `add_points(color = "red", size = 3)`
2. **Plot theme override**: `plot(theme = list(palette = "plasma"))`
3. **Instance theme**: `vis$set_theme(vistool_theme(...))`
4. **Package default**: `options(vistool.theme = vistool_theme(...))`

### Creating and using themes

```{r, eval=FALSE}
my_theme = vistool_theme(
  palette = "plasma",        # Color palette: "viridis", "plasma", "grayscale"
  text_size = 14,            # Base text size
  theme = "bw",              # ggplot2 theme: "minimal", "bw", "classic", etc.
  alpha = 0.7,               # Default transparency
  line_width = 1.5,          # Default line width
  point_size = 3,            # Default point size
  legend_position = "top",   # Legend position
  show_grid = TRUE,          # Show grid lines
  grid_color = "gray95",   # Grid color
  background = "white"     # Background color
)

obj = obj("TF_branin")
vis = as_visualizer(obj, type = "2d")

# setting the theme
vis$set_theme(my_theme)

# adding layers
vis$add_contours(color = "auto")

# override the theme for specific plots
vis$plot(theme = list(text_size = 12, alpha = 0.9))
```

## Layer-specific customization

All `add_*()` methods support style parameters that override theme defaults (see the reference!).
Below are common examples:

### `add_points()`
```{r, eval=FALSE}
vis$add_points(
  points,                   # Point data
  color = "auto",           # "auto" uses theme palette, or specify color
  size = NULL,              # Point size (NULL uses theme default)
  shape = 19,               # Point shape
  alpha = NULL,             # Transparency (NULL uses theme default)
  annotations = NULL,       # Text labels for points
  annotation_size = NULL,   # Size of annotations
  ordered = FALSE,          # Draw arrows between consecutive points
  arrow_color = NULL,       # Color of arrows
  arrow_size = 0.3          # Size of arrows
)
```

### `add_optimization_trace()`
```{r, eval=FALSE}
vis$add_optimization_trace(
  optimizer,                # Optimizer object with trace data
  line_color = NULL,        # Color of the trace line (NULL -> "auto" resolved at plot time)
  line_width = NULL,        # Width of trace line (NULL -> theme$line_width)
  line_type = "solid",      # Line type: "solid", "dashed", "dotted"
  npoints = NULL,           # Number of points to show from the trace (NULL -> all)
  npmax = NULL,             # Maximum number of points to show (post-processing cap)
  name = NULL,              # Name for the trace (legend)
  add_marker_at = 1,        # Iteration numbers where to add markers
  marker_size = 3,          # Size of markers
  marker_shape = 16,        # Marker shape (ggplot2 integer or plotly symbol)
  marker_color = NULL,      # Color of markers (NULL -> uses line color)
  show_start_end = TRUE,    # Highlight start/end points (2D only)
  alpha = NULL              # Alpha transparency (NULL -> theme$alpha)
)
```

### `add_annotation()`
`add_annotation()` attaches text callouts to any visualizer. Supply absolute coordinates via `x`, `y`, and `z` (surface plots), or pass `position = list(...)` with values in `[0, 1]` for relative placement. The method honors the current theme, resolving `color = "auto"`, default text size, and opacity automatically; you can still override them per annotation. Set `latex = TRUE` to typeset the label (requires `latex2exp` for ggplot2).

```{r, eval=FALSE}
# Absolute placement in data space
vis$add_annotation(
  text = "Optimum",
  x = 0, y = 0,
  color = "auto",
  opacity = 0.8
)

# Relative placement in the plot window
vis$add_annotation(
  text = "${}\\nabla f = 0$",
  latex = TRUE,
  position = list(x = 0.85, y = 0.2, reference = "figure"),
  color = "auto"
)
```

## Plot-level customization

The `plot()` method accepts theme overrides and functional parameters:

```{r, eval=FALSE}
vis$plot(
  # Theme override (merged with instance theme)
  theme = list(
    palette = "grayscale",      # Override palette for this plot
    text_size = 16,             # Override text size
    alpha = 0.5,                # Override default alpha
    theme = "classic"           # Override ggplot2 theme
  ),

  show_title = TRUE,            # Show/hide title
  plot_title = "My Plot",       # Custom title
  plot_subtitle = "Subtitle",   # Add subtitle
  show_legend = TRUE,           # Show/hide legend
  legend_title = "Legend",      # Custom legend title
  x_lab = "X Variable",         # Custom axis labels
  y_lab = "Y Variable",
  z_lab = "Z Variable",         # For surface plots
  x_limits = c(-2, 2),          # Custom axis limits
  y_limits = c(-1, 1),
  z_limits = c(0, 100)          # For surface plots
)
```

Note that you can further customize the returned plot objects directly by adding *manual layers*.
For some examples, see the [Advanced visualization](advanced_visualization.html) vignette.

## Automatic color management

All visual elements support automatic color assignment using `color = "auto"`:

- Colors are automatically assigned from the effective theme palette
- Each visualizer tracks its color index to ensure distinct colors for multiple layers
- The color index resets on each `plot()` call
- Manual colors override automatic assignment

```{r, eval=FALSE}
# Example: Multiple traces with auto colors
vis = as_visualizer(obj("TF_branin"), type = "2d")
vis$set_theme(vistool_theme(palette = "plasma"))

vis$add_optimization_trace(optimizer1, line_color = "auto")  # Gets first color
vis$add_optimization_trace(optimizer2, line_color = "auto")  # Gets second color
vis$add_points(custom_points, color = "auto")                # Gets third color

# Plot and colors will be consistent with plasma palette
p = vis$plot()
```

## Example: complete workflow

```{r, out.width='100%', out.height='700px'}
# 1. Create visualizer (computational setup only)
task = tsk("mtcars")$select(c("wt", "hp"))
learner = lrn("regr.lm")
vis = as_visualizer(task, learner = learner, n_points = 50, padding = 0.1)

# 2. Set instance theme
vis$set_theme(vistool_theme(
  palette = "plasma",
  text_size = 12,
  theme = "bw",
  alpha = 0.8
))

# 3. Add layers with automatic styling
vis$add_training_data(color = "auto")  # Uses theme palette
vis$add_boundary(color = "black")      # Manual override

# 4. Plot with theme override for this render only
vis$plot(theme = list(text_size = 14, legend_position = "top"), show_title = FALSE)
```

```{r, eval=FALSE}
# 5. Save uses the cached plot for efficiency
vis$save("my_plot.png", width = 8, height = 6)
```

## Package-level defaults

Set global defaults that apply to all new visualizers:

```{r, eval=FALSE}
# Set package default theme
options(vistool.theme = vistool_theme(
  palette = "plasma",
  text_size = 12,
  theme = "bw"
))

# All new visualizers will use this theme unless overridden
vis1 = as_visualizer(obj("TF_branin"))  # Uses plasma palette
vis2 = as_visualizer(obj("TF_gaussian1"))  # Also uses plasma palette

# Instance themes still override package defaults
vis1$set_theme(vistool_theme(palette = "grayscale"))  # Now uses grayscale
```

## See also

- [Loss functions](loss_functions.html)
- [Model predictions](model.html)
- [Objective functions](objective.html)
- [Optimization & traces](optimization_traces.html)
- [Advanced visualization](advanced_visualization.html)
