---
title: "Visualization of Loss Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization of Loss Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Loss Functions

```{r, message=FALSE}
library(vistool)
```

The `VisualizerLossFunction` class visualizes loss functions.
Loss functions are wrapped in `LossFunction` objects.
The package includes a few predefined loss functions that are used for classification and regression.

```{r}
as.data.table(dict_loss)
```

To get a loss function from the dictionary, use the `lss()` function
We retrieve the hinge loss function for classification.

```{r}
loss_function = lss("hinge")
```

# Visualization

To visualize the loss function, use the `as_visualizer()` function.
The `y_pred` arguments contains $y * f(x)$ and `y_true` the true value $y$.

```{r}
vis = as_visualizer(loss_function, y_pred = seq(-4, 4), y_true = 1)
```

Use the `plot()` method to plot the loss function.

```{r, out.width='100%'}
vis$plot()
```

# Regression Loss Functions

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%', out.height='1000px'}
library(mlr3misc)
library(plotly)
library(ggsci)

visualizers = pmap(list(c("huber", "l1_ae", "l2_se", "log-cosh"), pal_npg("nrc")(4)), function(key, color) {
  vis = as_visualizer(lss(key), y_pred = seq(-10, 10), y_true = 0)
  vis$init_layer_lines(color = color)
  vis$plot()
})

subplot(visualizers, nrows = 4, shareX = TRUE, shareY = TRUE) %>%
  layout(
    title = "Regression Loss Functions",
    xaxis = list(title = "y - f(x)"),
    yaxis = list(title = "Loss"))
```

# Classification Loss Functions

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%', out.height='00px'}
visualizers = pmap(list(c("cross-entropy", "hinge"), pal_npg("nrc")(2)), function(key, color) {
  vis = as_visualizer(lss(key), y_pred = seq(-4, 4), y_true = 1)
  vis$init_layer_lines(color = color)
  vis$plot()
})

subplot(visualizers, nrows = 2, shareX = TRUE, shareY = TRUE) %>%
  layout(
    title = "Regression Loss Functions",
    xaxis = list(title = "y - f(x)"),
    yaxis = list(title = "Loss"))
```
