---
title: "Visualization of Model Predictions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization of Model Predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE}
library(vistool)
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
```

The `Visualizer1DModel` and `Visualizer2DModel` class visualize the predictions of a model on tasks with 1 or 2 features.
Let's start with the `boston_housing` data set.
The goal is to predict the median value of owner-occupied homes in $1000s.
We subset the data set to only use the features `age` and `rm`.
The `age` feature is the proportion of owner-occupied units built prior to 1940 and `rm` is the average number of rooms per dwelling.

```{r}
task = tsk("boston_housing")
task$select(c("age", "rm"))
```

We load the support vector machine learner for regression.

```{r}
learner = lrn("regr.svm")
```

Now we create a visualizer object with the `as_visualizer()` function.

```{r}
vis = as_visualizer(task, learner)
```

First, the learner is trained on the entire task.
After that a grid is created for the two features and the predictions of the model are computed for each grid point.
The predictions are then visualized using a 3D scatter plot.

```{r, out.width='100%', out.height='700px'}
vis$plot()
```

We can also create a 2D contour plot for the predictions.

```{r, out.width='100%', out.height='700px'}
vis$init_layer_contour()
vis$plot()
```

To switch back to the 3D scatter plot initialize the scatter layer again.

```{r}
vis$init_layer_surface()
```

We can add the training points to the plot.

```{r, out.width='100%', out.height='700px'}
vis$add_training_data()
vis$plot()
```

It is also possible to visualize classification tasks.
We use the `pima` data set and impute the missing values.
We select the features `insulin` and `mass` and train a support vector machine for classification.

```{r}
task = tsk("pima")
task = po("imputemean")$train(list(task))[[1]]
task$select(c("insulin", "mass"))
learner = lrn("classif.svm", predict_type = "prob")
```

We create a visualizer object and plot the predictions.
The z-axis shows the probability of the positive class.

```{r, out.width='100%', out.height='700px'}
vis = as_visualizer(task, learner)
vis$plot()
```

We can add the decision boundary to the plot.

```{r, out.width='100%', out.height='700px'}
vis$add_decision_boundary()
vis$plot()
```
