---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  dev = "png",
  out.width = "100%",
  fig.retina = 2
)
```

```{r, message=FALSE}
library(vistool)
set.seed(1)
```

## What is vistool?

`vistool` helps visualize optimization concepts and model predictions:
- Loss functions (regression/classification)
- Objective functions (1D/2D test functions and custom objectives)
- Optimization traces (GD, Momentum, Nesterov)
- Model prediction surfaces and decision boundaries

You can render either with `ggplot2` (2D) or interactively with `plotly` (3D surfaces).

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("slds-lmu/vistool")
```

### Plotly backend and static export

- Interactive surfaces use `plotly`.
- Saving plotly surfaces as static images uses `plotly::save_image()`, which relies on the Python package `kaleido` (a separate Chrome/Chromium installation is required).
- On first use of save/export, `vistool` triggers `reticulate::py_require("kaleido")` to provision a cached Python environment automatically.

If you prefer to manage Python yourself, ensure `kaleido` is installed and that `reticulate` points to your Python:

```r
# optional manual setup
install.packages("reticulate")
# Sys.setenv(RETICULATE_PYTHON = "/path/to/python")
reticulate::py_install("kaleido")
```

## Quick tour

TODO: explain workflow: 1) `as_visualizer`, 2) adding layers, 3) plot, 4) save; theme system

TODO: visualize ...

- [Loss Functions](loss_functions.html)
- [Model Predictions](model.html)
- [Objective Functions](objective.html)
- [Optimization & Traces](optimization_traces.html)

For customization options, see:
- [Customization Guide](customization_guide.html)
- [Advanced Visualization](advanced_visualization.html)
